Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-05-09 article-models-ttderived.tex
extension algebras finished (initial version)

>>>>

\subsection{Extension algebras}
Let $P$ be a family over an extended context $\ctxext{\Gamma}{A}$. We could
mimic extension by requiring to have terms
\begin{align*}
\jhom*{\Gamma}{{A}{P}}{A}{e}\\
\jhom*{{\Gamma}{A}}{{P}{\jcomp{}{e}{P}}}{P}{f}.
\end{align*}
We say that a quadruple $(A,P,e,f)$ as above is an extension algebra if the
following diagrams commute judgmentally:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\ctxext{A}{{P}{\jcomp{}{e}{P}}} 
  \ar{d}[swap]{\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}
    %\tmext{\ctxwk{\jcomp{}{e}{P}}{e}}{\idtm{\jcomp{}{e}{P}}}
    } 
  \ar{r}{\jvcomp{}{\idtm{A}}{f}
    %\tmext{\ctxwk{\ctxext{P}{\jcomp{}{e}{P}}}{\idtm{A}}}{f}
    } 
  & \ctxext{A}{P} \ar{d}{e}\\
\ctxext{A}{P} \ar{r}[swap]{e} & A
\end{tikzcd}
\end{equation*}
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\ctxext{P}{{\jcomp{}{e}{P}}{\jcomp{}{\bar{e}}{{}{e}{P}}}} 
  \ar{r}{\jvcomp{}{\idtm{P}}{\jcomp{}{e}{f}}}
  \ar{d}[swap]{
    \jvcomp{}{f}{\idtm{\jcomp{}{\bar{e}}{{}{e}{P}}}}
%    \tmext
%      {\ctxwk{\jcomp{}{\bar{e}}{{}{e}{P}}}{f}}
%      {\idtm{\jcomp{}{\bar{e}}{{}{e}{P}}}}
    }
& \ctxext{P}{\jcomp{}{e}{P}} \ar{d}{f}\\
\ctxext{P}{\jcomp{}{e}{P}} \ar{r}[swap]{f} & P
\end{tikzcd}
\end{equation*}
where we have taken 
$\bar{e}\defeq\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}$, which
is the morphism on the left of the first diagram (taking the upper morphism
would yield a judgmentally equal result). To see that the above diagram makes
sense, note that we can compute the upper left family in two different ways.
First, we have the judgmental equalities
\begin{align*}
\jcomp{}{\bar{e}}{\jcomp{}{e}{P}}
& \jdeq 
  \jcomp{}{\jcomp{}{\bar{e}}{e}}{P}
  \\
& \jdeq
  \jcomp{}{
    \jcomp{}{\jvcomp{}{f}{\idtm{\jcomp{}{\bar{e}}{{}{e}{P}}}}}{e}}{P}
  \\
& \jdeq
  \jcomp{}{\jvcomp{}{f}{\idtm{\jcomp{}{\bar{e}}{{}{e}{P}}}}}{
    \jcomp{}{e}{P}}
  \\
& \jdeq
  \subst
    {\tmext{\ctxwk{\ctxext{P}{\jcomp{}{e}{P}}}{\idtm{A}}}{f}}
    {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \subst
    {f}
    { {\ctxwk{\ctxext{P}{\jcomp{}{e}{P}}}{\idtm{A}}}
      {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}
      }
  \\
& \jdeq
  \subst
    {f}
    {\ctxwk{\ctxext{P}{\jcomp{}{e}{P}}}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \jcomp{}{f}{\jcomp{}{e}{P}}
\end{align*}
so the downwards arrow on the left indeed makes sense. Second, we have the
judgmental equalities
\begin{align*}
\ctxext{\jcomp{}{e}{P}}{\jcomp{}{\bar{e}}{\jcomp{}{e}{P}}}
& \jdeq
  \ctxext{\jcomp{}{e}{P}}{\subst{\jvcomp{}{e}{\idtm{\jcomp{}{e}{P}}}}
    {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}}
  \\
& \jdeq
  \ctxext{\jcomp{}{e}{P}}{\subst
    {\idtm{\jcomp{}{e}{P}}}
    {{\ctxwk{\jcomp{}{e}{P}}{e}}
    {\ctxwk{\ctxext{A}{{P}{\jcomp{}{e}{P}}}}{\jcomp{}{e}{P}}}}}
  \\
& \jdeq
  \ctxext{\jcomp{}{e}{P}}{\subst
    {\idtm{\jcomp{}{e}{P}}}
    {{\ctxwk{\jcomp{}{e}{P}}{e}}
    {\ctxwk{\ctxext{{A}{P}}{\jcomp{}{e}{P}}}{\jcomp{}{e}{P}}}}}
  \\
& \jdeq
  \ctxext{\jcomp{}{e}{P}}{\subst
    {\idtm{\jcomp{}{e}{P}}}
    {{\ctxwk{\jcomp{}{e}{P}}{e}}
    {\ctxwk{\jcomp{}{e}{P}}{{\ctxext{A}{P}}{\jcomp{}{e}{P}}}}}}
  \\
& \jdeq
  \ctxext{\jcomp{}{e}{P}}{\subst
    {\idtm{\jcomp{}{e}{P}}}
    {\ctxwk{\jcomp{}{e}{P}}{\subst{e}
    {\ctxwk{\ctxext{A}{P}}{\jcomp{}{e}{P}}}}}}
  \\
& \jdeq
  \ctxext{\jcomp{}{e}{P}}{\subst{e}
    {\ctxwk{\ctxext{A}{P}}{\jcomp{}{e}{P}}}}
  \\
& \jdeq
  \ctxext{\jcomp{}{e}{P}}{\jcomp{}{e}{\jcomp{}{e}{P}}}
  \\
& \jdeq
  \jcomp{}{e}{\ctxext{P}{\jcomp{}{e}{P}}}
\end{align*}
and we indeed have the morphism $\jcomp{}{e}{f}$ from
$\jcomp{}{e}{\ctxext{P}{\jcomp{}{e}{P}}}$ to $\jcomp{}{e}{P}$.
\subsection{Extension-empty algebras}
\begin{defn}
Let $P$ be a family over the extended context $\ctxext{\Gamma}{A}$, let
$\jterm{\Gamma}{A}{x}$ and $\jterm{{\Gamma}{A}}{P}{f}$ be terms. Then the
quadruple $(A,P,x,f)$ is said to be an \emph{empty algebra in context $\Gamma$}
if the following judgmental equalities hold:
\begin{align*}
\jfameq*{\Gamma}{\subst{x}{P}}{A}\\
\jtermeq*{\Gamma}{A}{\subst{x}{f}}{x}.
\end{align*}
\end{defn}
